Deciphering the Trends: A Comprehensive Analysis of US Birth Rates and Educational Attainment from 1980 to 2020

### 2024-02-10 outputs/paper/paper.qmd
paper.qmd edits

This paper analyses the changing landscape of birth rates in the United States over forty years, focusing on its interrelationships among age, education level, and marital status. Using extensive data from the American Economics Association, we employ rigorous statistical analysis to uncover the nuances of these trends. Our findings estimated the impact of  socioeconomic traits, such as education level and marital status, on fertility decisions to explain decrease in fertility seen over time.Particularly highlighting the shift towards later-life parenthood and the changing dynamics of family formation when it comes to marriage. The analysis shows how changes in these demographic factors have shaped fertility patterns.

Our analysis, conducted using the R programming language [@citeR], reveals significant shifts in birth patterns that have far-reaching implications. The larger context of demographic shifts and their implications for societal structures, economic planning, and policy formulation is the driving force behind this study. By breaking down the trends by demographics, we hope to close a critical gap in our understanding of the interactions between socioeconomic factors and fertility decisions.

The dataset was obtained from the data section for “The Puzzle of Falling US Birth Rates since the Great Recession” [@citepaper], a paper from the Journal of Economic Perspectives [@citewebsite]. Analysis of the data was done using the R programming language [@citeR] in conjunction with the following packages: tidyverse [@tidyverse] and Knitr [@knitr]. The figures were created using ggplot2 [@citeggplot]. We began by constructing a comprehensive line graph illustrating the overarching trend of the birth rate per 1000 women across all women aged 15-44 in the United States from 1980-2020. We then plotted three-line graphs to analyze the patterns in birth rate with relation to three distinct demographic groups defined by marital status, educational level, and age group of the mother. This granular breakdown facilitated a more nuanced understanding of how birth rates varied within these demographic cohorts for the specified period. This helped us study the overall trend concerning these three demographic groups.

@@

The American Economics Association [@citewebsite] has made available a dataset that provides a thorough analysis of birth rates over a long period, broken down by a variety of demographic variables. Among the detailed records of birth rates are those by age group, education level, and marital status. Birth rates by age group and marital status are gathered from CDC Vital Statistics Births Reports. Birth rates by mother’s education is calculated by the using  National Center for Health Statistics (NCHS) Vital Statistics birth microdata, SEER population data, and the Current Population Survey. The age group data covers the years 1980 to 2020 and is particularly useful for understanding how birth rates have changed over time across different age brackets. The education level data covers the years 1992 to 2019 and gives insights into how educational attainment influences birth rates. The marital status data, which covers the years 1980 to 2019, provides insight into the variations in birth rates between married and single people, reflecting changes in society and individual preferences. For this paper, there were no suitable alternative data sources.

@@

Age is one reliable variables in this study because age group measurement is less ambiguous and more direct. It is determined using official documentation, such as birth certificates, passports, or ID cards. This process ensures high accuracy in categorizing individuals into their respective age groups. However, the accuracy and availability of official records can vary, which can affect the precision of age group data. Fortunately any inaccuracy due to this is heavily mitigated by the grouping of age groups rather than if the analysis was done on individual ages.

@@

The Mother's Education item on the U.S. Standard Certificate of Live Birth is used to determine the mother's educational attainment. This information is provided by the mother or a surrogate and asks for the highest degree or level of education completed at the time of delivery. While this method allows for a direct report of educational attainment, it suffers from some information biases, such as social desirability bias [@citebias]which may skew the results towards higher education levels. Further self reported answers may be influenced by the individual's understanding and interpretation of the question.

@@

@fig-2 depicts the fluctuations in birth rates for 6 different age groups between 1980 – 2020. Age groups 15-19, 20-24 and 25-29 saw a sharp decline in birth rate over the four decades with 15-19 experiencing the steepest decline. The birth rate for this age group went from 53 births per 1000 women in 1980 to 15.3 births in 2020. Birth rates for women in their 20s also saw noticeable declines, though the decrease was less sharp compared to the teenage group. Specifically, the birth rate for women aged 20-24 fell from 115.1 per 1,000 in 1980 to 62.8 per 1,000 in 2020, while for women aged 25-29, it declined from 112.9 per 1,000 to 90.0 per 1,000 over the same period.

@@

The analysis shows a complex relationship between education levels and birth rates, highlighting how fertility decisions are influenced by educational attainment. The sharp decline in birth rates among the most and least educated women indicates that family size preferences are strongly influenced by education. For higher educated women, the tendency towards lower birth rates further reflects the importance placed on career and financial stability, with education offering both the means and the justification for delaying or limiting childbearing [@ncbi].

>>>>

This paper analyzes the changing landscape of birth rates in the United States over forty years, focusing on its interrelationships among age, education level, and marital status. Using extensive data from the American Economics Association, we employ statistical analysis to uncover the nuances of these trends. Our findings estimated the impact of  socioeconomic traits, such as education level and marital status, on fertility decisions to explain the decrease in fertility seen over time. They particularly highlight the shift towards later-life parenthood and the changing dynamics of family formation when it comes to marriage. The analysis shows how changes in these demographic factors have shaped fertility patterns.

Our analysis, conducted using the R programming language [@citeR], reveals significant shifts in birth patterns that have far-reaching implications. The larger context of demographic shifts and their implications for societal structures, economic planning, and policy formulation is the driving force behind this study. By breaking down the trends by demographic, we hope to close a critical gap in our understanding of the interactions between socioeconomic factors and fertility decisions.

The dataset was obtained from the data section for “The Puzzle of Falling US Birth Rates since the Great Recession” [@citepaper], a paper from the Journal of Economic Perspectives [@citewebsite]. Analysis of the data was done using the R programming language [@citeR] in conjunction with the following packages: tidyverse [@tidyverse] and Knitr [@knitr]. The figures were created using ggplot2 [@citeggplot]. We began by constructing a comprehensive line graph illustrating the overarching trend of the birth rate per 1000 women across all women aged 15-44 in the United States from 1980-2020. We then plotted three-line graphs to analyze the patterns in birth rate with relation to three distinct demographic groups defined by marital status, educational level, and age group of the mother. This breakdown facilitated a more nuanced understanding of how birth rates varied within these demographic cohorts for the specified period. This helped us study the overall trend concerning these three demographic groups.

@@

The American Economics Association [@citewebsite] has made available a dataset that provides a thorough analysis of birth rates over a long period, broken down by a variety of demographic variables. Among the detailed records of birth rates are those by age group, education level, and marital status. Birth rates by age group and marital status are gathered from CDC Vital Statistics Births Reports. Birth rates by mother’s education is calculated by the  National Center for Health Statistics (NCHS) Vital Statistics birth microdata, SEER population data, and the Current Population Survey. The age group data covers the years 1980 to 2020 and is particularly useful for understanding how birth rates have changed over time across different age brackets. The education level data covers the years 1992 to 2019 and gives insights into how educational attainment influences birth rates. The marital status data, which covers the years 1980 to 2019, provides insight into the variations in birth rates between married and single people, reflecting changes in society and individual preferences. For this paper, there were no suitable alternative data sources.

@@

Age is one reliable variable in this study because age group measurement is less ambiguous and more direct. It is determined using official documentation, such as birth certificates, passports, or ID cards. This process ensures high accuracy in categorizing individuals into their respective age groups. However, the accuracy and availability of official records can vary, which can affect the precision of age group data. Fortunately, any such inaccuracy is heavily mitigated by the grouping of age groups rather than performing the analysis on individual ages.

@@

The Mother's Education item on the U.S. Standard Certificate of Live Birth is used to determine the mother's educational attainment. This information is provided by the mother or a surrogate and asks for the highest degree or level of education completed at the time of delivery. While this method allows for a direct report of educational attainment, it suffers from some information biases, such as social desirability bias [@citebias] which may skew the results towards higher education levels. Further self-reported answers may be influenced by the individual's understanding and interpretation of the question.

@@

@fig-2 depicts the fluctuations in birth rates for 6 different age groups between 1980 – 2020. Age groups 15-19, 20-24 and 25-29 saw a sharp decline in birth rate over the four decades with 15-19 experiencing the steepest decline. The birth rate for this age group went from 53 births per 1000 women in 1980 to 15.3 births in 2020. Birth rates for women in their 20s also saw noticeable declines, though the decrease was less sharp compared to the teenaged group. Specifically, the birth rate for women aged 20-24 fell from 115.1 per 1,000 in 1980 to 62.8 per 1,000 in 2020, while for women aged 25-29, it declined from 112.9 per 1,000 to 90.0 per 1,000 over the same period.

@@

The analysis shows a complex relationship between education levels and birth rates, highlighting how fertility decisions are influenced by educational attainment. The sharp decline in birth rates among the most and least educated women indicates that family size preferences are strongly influenced by education. For women with higher levels of education, the tendency towards lower birth rates further reflects the importance placed on career and financial stability, with education offering both the means and the justification for delaying or limiting childbearing [@ncbi].
